Is high social class always beneficial for survival? Long-term perspectives of social inequality in mortality among adults and elderly: Northern Sweden 1851--2013

### 2017-07-24 article.Rmd
First update article version

---
title: Long-term perspectives of social inequality in mortality among adults and elderly
  in northern Sweden 1851--2013
author: "Sören Edvinsson and Göran Broström"
date: "28 juni 2017"
output:
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::pdf_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
link-citations: yes
bibliography: mybib.bib
biblio-style: apalike
---

@@

One of the central aspects of survival is social class and access to economic
and other resources. Even in present-day welfare societies, social position is a
strong determinant when it comes to health and mortality and the impact even
seems to be increasing [@kuetal04, @frlu07]. It has been suggested that
"... social conditions have been, are and will continue to be irreducible
determinants of health outcomes and therefore deserves their appellation as
'fundamental causes' of disease and death" [@liph95]. For a long time,
the general view has been that socio-economic health inequalities were high in
historical societies, probably higher than in modern societies. This is a
reasonable assumption since these societies were in most cases characterised by
very large differences between groups and classes. Knowing that access to
resources provides advantages in all ways, the health advantage of higher
classes ought to be obvious. @ant67 and @smi83 suggest that social
inequality in health has passed through different historical phases. According
to them, social health differences were comparatively small during the
pre-transitional phase when space was a strong determinant for the spread of
disease. Differences then increased during the transitional phase when mortality
declined and wealthy groups used their resources to gain better health. Finally,
health differences decreased again in modern low-mortality settings when instead
health-related behaviour became important, resulting in marginal differences.
@omran82 states, in the third proposition in his theory of the
epidemiological transition, that even if the class differential in mortality was
maintained during the transition, the decline set in earlier and was faster
among privileged groups.

>>>>

---
title: Is high social class always beneficial for survival? Long-term perspectives of social inequality in mortality among adults and elderly: Northern Sweden 1851--2013
author: "Sören Edvinsson and Göran Broström"
date: "24 JulY 2017"
output:
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::pdf_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
link-citations: yes
bibliography: mybib.bib
biblio-style: apalike
---

@@

One of the central aspects of survival is social class and access to economic
and other resources. Even in present-day welfare societies, social position is a
strong determinant when it comes to health and mortality and the impact even
seems to be increasing [@kuetal04, @macketal16, @frlu07, @broba12, @strand2010]. It has been suggested that
"... social conditions have been, are and will continue to be irreducible
determinants of health outcomes and therefore deserves their appellation as
'fundamental causes' of disease and death" [@liph95]. For a long time,
the general view has been that socio-economic health inequalities were high in
historical societies, probably higher than in modern societies. This is a
reasonable assumption since these societies were in most cases characterised by
very large differences between groups and classes. Knowing that access to
resources provides advantages in all ways, the health advantage of higher
classes ought to be obvious. @ant67 and @smi83 suggest that social
inequality in health has passed through different historical phases. According
to them, social health differences were comparatively small during the
pre-transitional phase when space was a strong determinant for the spread of
disease. Differences then increased during the transitional phase when mortality
declined and wealthy groups used their resources to gain better health. Finally,
health differences decreased again in modern low-mortality settings when instead
health-related behaviour became important, resulting in marginal differences.
@omran82 states, in the third proposition in his theory of the
epidemiological transition, that even if the class differential in mortality was
maintained during the transition, the decline set in earlier and was faster
among privileged groups.
